---
title       : Base R Graphs/Charts
subtitle    : 
author      : David Boose
job         : 
framework   : bootstrap        ## {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  ## {highlight.js, prettify, highlight}
hitheme     : tomorrow      ## 
widgets     : []            ## {mathjax, quiz, bootstrap}
mode        : standalone ## {standalone, draft}
---

### Base R Version

#### Data
Let us begin by simulating our sample data of 3 factor variables and 4 numeric variables.

```{r echo = T}
## Simulate some data

## 3 Factor Variables
FacVar1=as.factor(rep(c("level1","level2"),25))
FacVar2=as.factor(rep(c("levelA","levelB","levelC"),17)[-51])
FacVar3=as.factor(rep(c("levelI","levelII","levelIII","levelIV"),13)[-c(51:52)])

## 4 Numeric Vars
set.seed(123)
NumVar1=round(rnorm(n=50,mean=1000,sd=50),digits=2) ## Normal distribution
set.seed(123)
NumVar2=round(runif(n=50,min=500,max=1500),digits=2) ## Uniform distribution
set.seed(123)
NumVar3=round(rexp(n=50,rate=.001)) ## Exponential distribution
NumVar4=2001:2050

simData=data.frame(FacVar1,FacVar2,FacVar3,NumVar1,NumVar2,NumVar3,NumVar4)
```
These commands create the seven different variables in the data set. FacVar1, 2, and 3 are the qualitative factor variables. They are created by using the "c()" command to read in the possible variable values, and the "rep()" command to repeat that process enough times to generate 50 values of each variable. For FacVar2 and FacVar3, one or two values need to be removed from the data set to bring it to a total of 50 observations. The command "as.factor()" ensures that the variables are declared as factor variables.

For the numeric variables, four different processes are used to generate values. For NumVar1, 50 values are drawn at random from a normal distribution with mean 1000 and standard deviation 50. NumVar2 is generated by drawing 50 vaules from a uniform distribution ranging between 500 and 1000. Both NumVar1 and NumVar2 are rounded to two decimal places using the "round()" command. NumVar3 consists of 50 values chosen from an exponential distribution with a rate parameter 0.001. These values are rounded to integers. Finally, NumVar4 is simply a list of sequential numbers from 2001 - 2050, inclusive.

Once the variable vectors have been created, they are combined into the dataframe "simData" by defining the dataframe.

Now let's make some plots. We'll start by examining the distributions of the values of the different variables. For numeric variables, this is done using four types of graphs, which are shown for NumVar1 below:

#### One Variable: Numeric Variable

```{r echo = T, results="asis", warning=FALSE,message=FALSE}
plot(simData$NumVar1,type="o") ## Index plot
hist(simData$NumVar1) ## histogram
plot(density(simData$NumVar1)) ## Kernel density plot
boxplot(simData$NumVar1) ## box plot
```

The first graph is an index plot. The x-axis is the position of the value in the vector (1 - 50), and the y value is the value of NumVar1 at that position. This plot simply shows the values of the variable. the "type='o'" command allows two or more equivalent values to be plotted in on top of one another, but because each value is plotted according to its index number, this won't happen in this graph.
The other three plots show the distribution of the data points, but not the values themselves. The second graph is a histogram. The ranges of the bins for the histogram is determined using a default algorithm. The third plot is a Kernel density plot, which is something like a smoothed histogram. It uses an algorithm to compute the probability density of a population of values, based on the sample given. The plot then shows the probability density function. Finally, the fourth plot is a boxplot, which uses a rectangle to show the interquartile range of the data, a horisontal line in the box to indicate the median, and dashed vertical lines extending from the first and third quartiles to the ends of the range of the values. 

For the factor variables, the "distribution" is simpler. Because there are only 2, 3, or 4 values the variables can take, the distribution is just the number of times each value is represented in the dataset. Because we created the variables by repeatedly choosing the different values in turn, the levels have approximately equal frequencies. The distributions are perhaps best displayed in a table of frequencies, but the two graphs below can also be used:

#### One Variable: Factor Variable

```{r echo = T, results="asis", warning=FALSE,message=FALSE}
plot(simData$FacVar3) ## bar plot

## pie chart - Not the best graph --- use with caution
counts=table(simData$FacVar3) ## get counts
labs=paste(simData$FacVar3,counts)## create labels
pie(counts,labels=labs) ## plot
```

The bar plot displays the frequency of each level as the height of a bar. The pie chart shows the different levels as proportions contributing to a whole. 

#### Two Variables: Two Numeric Variables

Usually, we are interested in plotting two or more variables, to see the relationship between them. 
The graphs below show two of the distribution graphs from above--the index plot and the kernel density plot--with the data from two variables overlaid on one another. The scatterplot shows the relationship between the two variables by plotting them as x,y pairs.

```{r echo = T, results="asis", warning=FALSE,message=FALSE}
plot(simData$NumVar1,type="o",ylim=c(0,max(simData$NumVar1,simData$NumVar2)))## index plot with one variable
lines(simData$NumVar2,type="o",lty=2,col="red")## add another variable

## Let's draw density plots : https://stat.ethz.ch/pipermail/r-help/2006-August/111865.html
dv1=density(simData$NumVar1) 
dv2=density(simData$NumVar2)
plot(range(dv1$x, dv2$x),range(dv1$y, dv2$y), type = "n", xlab = "NumVar1(red) and NumVar2 (blue)",
     ylab = "Density")
lines(dv1, col = "red")
lines(dv2, col = "blue")

## scatterplots
plot(simData$NumVar1,simData$NumVar2)
```

In the code above, the axis ranges are given in a couple of different ways. For the index plot, the range of y values is from zero to the largest value found in either NumVar1 or NumVar2. The range of the x-axis will be determined by hte number of observations in the dataset. For the kernel density plots, the x and y axes are given by the function "range()", applied to the probability density calculations on the two variables. The "type = n" parameter means that the data points are not plotted; the "lines()" commands then add the lines to the plot. 

#### Two Variables: Two Factor Variables

When a dataset contains more than one factor variable, it is often of interest to see how often different levels of the variable co-occur. If certain combinations of values occur more often than would be expected by chance, we would say there is an association between the two variables. As with the single-variable distributions above, these patterns are often best represented in a table; but some graphical representations can be used as well.

```{r echo = T, results="asis", warning=FALSE,message=FALSE}
## Mosaic plot
plot(table(simData$FacVar2,simData$FacVar3))

## barplots
bartable=table(simData$FacVar2,simData$FacVar3) ## get the cross tab
barplot(bartable,beside=TRUE, legend=levels(unique(simData$FacVar2))) ## plot 
barplot(bartable, legend=levels(unique(simData$FacVar2))) ## stacked
barplot(prop.table(bartable,2)*100, legend=levels(unique(simData$FacVar2))) ## stacked 100%
```

The mosaic plot uses rectangles 
#### Two Variables: One Factor and One Numeric
```{r echo = T, results="asis", warning=FALSE,message=FALSE}
## Box plots for the numeric var over the levels of the factor var
plot(simData$FacVar1,simData$NumVar1) 

## density plot of numeric var across multiple levels of the factor var
level1=simData[simData$FacVar1=="level1",] 
level2=simData[simData$FacVar1=="level2",]

dv3=density(level1$NumVar1)
dv4=density(level2$NumVar1)

plot(range(dv3$x, dv4$x),range(dv3$y, dv4$y), type = "n", xlab = "NumVar1 at Level1 (red) and NumVar1 at Level2 (blue)",ylab = "Density")
lines(dv3, col = "red")
lines(dv4, col = "blue")

## Mean of one numeric var over levels of one factor var
meanagg=aggregate(simData$NumVar1, list(simData$FacVar3), mean)

dotchart(meanagg$x,labels=meanagg$Group.1) ## Dot Chart 
barplot(meanagg$x,names.arg=meanagg$Group.1)## Bar plot 
## Question: Is a bar plot even appropriate when displaying a mean--- a point?
```

#### Three Variables: Three Factor Variables

```{r echo = T, results="asis", warning=FALSE,message=FALSE}
par(mfrow=c(1,2))

bar1table=table(level1$FacVar2,level1$FacVar3)
barplot(bar1table,beside=TRUE, main="FacVar1=level1")

bar2table=table(level2$FacVar2,level2$FacVar3)
barplot(bar2table,beside=TRUE, main="FacVar1=level2", legend=levels(unique(level2$FacVar2)))
```

#### Three Variables: One Numeric and Two Factor Variables
```{r echo = T, results="asis", warning=FALSE,message=FALSE}
par(mfrow=c(1,1))
## boxplot of NumVar1 over an interaction of 6 levels of the combination of FacVar1 and FacVar2
boxplot(NumVar1~interaction(FacVar1,FacVar2),data=simData)

## Mean of 1 Numeric over levels of two factor vars
meanaggg=aggregate(simData$NumVar1, list(simData$FacVar1,simData$FacVar2), mean)
meanaggg=meanaggg[order(meanaggg$Group.1),]
meanaggg$color[meanaggg$Group.2=="levelA"] = "red"
meanaggg$color[meanaggg$Group.2=="levelB"] = "blue"
meanaggg$color[meanaggg$Group.2=="levelC"] = "darkgreen"  

dotchart(meanaggg$x,labels=meanaggg$Group.2, groups=meanaggg$Group.1,color=meanaggg$color) ## dotchart

interaction.plot(meanaggg$Group.2,meanaggg$Group.1,meanaggg$x,type="b", col=c(1:2),pch=c(18,24)) ## interaction plot - line plots of means

## some a bar plot
par(mfrow=c(1,2))

level1=meanaggg[meanaggg$Group.1=="level1",]
level2=meanaggg[meanaggg$Group.1=="level2",]

barplot(level1$x,names.arg=level1$Group.2, main="FacVar1=level1")
barplot(level2$x,names.arg=level2$Group.2, main="FacVar1=level2")
```

#### Three Variables: Two Numeric and One Factor Variables

```{r echo = T, results="asis", warning=FALSE,message=FALSE}
## Scatter plot with color identifying the factor variable
par(mfrow=c(1,1))
plot(simData$NumVar1,simData$NumVar2, col=simData$FacVar1)
legend("topright",levels(simData$FacVar1),fill=simData$FacVar1)
```

#### Three Variables: Three Numeric Variables 

```{r echo = T, results="asis", warning=FALSE,message=FALSE}
## NumVar4 is 2001 through 2050... possibly, a time variable - use that as the x-axis
plot(simData$NumVar4,simData$NumVar1,type="o",ylim=c(0,max(simData$NumVar1,simData$NumVar2)))## join dots with lines

lines(simData$NumVar4,simData$NumVar2,type="o",lty=2,col="red")## add another line

## Bubble plot - scatter plot of NumVar1 and NumVar2 with individual observations sized by NumVar3
# http://flowingdata.com/2010/11/23/how-to-make-bubble-charts/
radius <- sqrt( simData$NumVar3/ pi )
symbols(simData$NumVar1,simData$NumVar2,circles=radius, inches=.25,fg="white", bg="red", main="Sized by NumVar3")
```

#### Scatterplot Matrix of all Numeric Vars, colored by a Factor variable
```{r echo = T, results="asis", warning=FALSE,message=FALSE}
pairs(simData[,4:7], col=simData$FacVar1)
```

#### References

Besides the link from flowingdata.com referred to in the context of the bubble plot, additional websites were used as references. 
http://www.harding.edu/fmccown/r/
http://www.statmethods.net/
